\begin{abstract}
We present a reproducible computational framework---with Lean~4 mechanized proofs and deterministic Python simulations---that treats multi-scale model reduction as a closure-audit calculus: each ``physics theory'' is a packaging operator built from a lens (retained macro information), a completion (canonical reconstruction), and diagnostics that certify when the reduced description is self-consistent. We instantiate the Six-Birds lens/completion/closure dictionary as a multi-scale layer template, centering three cross-domain certificates: closure coherence (idempotence), audit monotonicity (data-processing inequality), and route mismatch (noncommutation of coarse-graining with dynamics). We operationalize this calculus via a reproducible workflow (Python artifacts + Lean-certified lemmas) and demonstrate it on four instantiations: quantum dephasing, discrete BGK-like kinetic closure, filtered Burgers (LES), and nonlinear averaging (backreaction). Under null conditions (no coarse-graining, zero heterogeneity, or exact projections), idempotence defects and route mismatches are $\approx 0$; under controlled interventions they separate cleanly. Dephasing is exactly idempotent; sampled CPTP channels satisfy quantum relative-entropy DPI with no violations beyond tolerance; unitary evolution produces nonzero mismatch that grows with dephasing strength. In the BGK model, packaging coherence improves as collision strength increases, with explicit failure modes when the equilibrium family or retained moments are misspecified. In LES and averaging toys, mismatch grows with filter width or heterogeneity and yields canonical subgrid/backreaction correction terms. This reframes ``physics theories'' as comparable closure layers and localizes which effective terms are forced by a chosen scale and audit. We do not derive continuum theories or prove analytic bounds; we report Lean-certified algebraic facts and numerical certificates under stated controls.
\end{abstract}

\vspace{1ex}
\noindent\textbf{keywords:} model reduction; coarse-graining; closure operators; idempotence; route mismatch; audit metrics; data-processing inequality; information theory; formal verification; Lean; reproducible research
